An FPGA Implementation of a Long Short-Term Memory Neural Network

### 2016-07-25 paper/paper.tex
Final Version of the Text and Short Paper

%% Additional Packages
\usepackage{cite}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{siunitx}

%% Macros
\newcommand{\mb}[1]{\mathbf{#1}}

@@

\begin{abstract}
This work proposes a hardware architecture for a Long Short-Term Memory (LSTM) Neural Network, aiming
to outperform software implementations, by exploiting its inherent parallelism.
The main design decisions are presented, along with the proposed network architecture. A description of the main
building blocks of the network is also presented. The network is synthesized for various sizes and platforms,
and the performance results are presented and analysed. The synthesized network achieves a 251 times speed-up
over a custom-built software network, proving the benefits of parallel computation for this kind of network.
\end{abstract}

\section{Introduction}\label{sec:intro}
\IEEEPARstart{N}{eural} Networks are one of the most commonly used techniques in Deep Learning. This
particular type of network, LSTM, is a recursive network, given that the neuron outputs in a certain
time step are also fed as the input in the next time step, and that their memory elements can make
sense of patterns within data sequences, unlike classical recursive neural networks.
These algorithms have been profusely implemented in software, and their practical applications are plentiful.
However, the benefits of the inherent parallelism offered by a dedicated hardware
platform are not availed, and there are relatively few implementations of Machine Learning algorithms in
these kind of platforms.

Hardware platforms would achieve a considerable speed-up over software implementations, which could prove
useful for high data throughput systems, where the calculation overhead is critical and limits the performance of the systems.
Furthermore, a hardware implementation can even benefit \emph{offline} Deep Learning tasks by providing the results
of a given experiment faster than in a regular CPU, yielding an increase in scientific productivity.

Hitherto, there is only one implementation~\cite{Chang15}, but its performance is undermined
by the external memory access. This implementation aims to make use of internal FPGA memory resources,
and therefore achieve a higher speed performance.

In overview, this article begins by explaining what are LSTM networks -- and their mathematical
details -- in Section~\ref{sec:lstmnn}, followed by providing a quick overview of their state
of the art in Section~\ref{sec:relwork}. Section~\ref{sec:proparch} outlines the proposed
hardware architecture and its main constituent modules, and the performance and synthesis
results are reported in Section~\ref{sec:results}. A final concluding remark is given in
Section~\ref{sec:concl}.

\section{LSTM Neural Networks}\label{sec:lstmnn}
LSTM Networks were originally formulated in~\cite{Hoch97}, but its formulation has been incrementally
updated in~\cite{Gers00} and~\cite{Gers2000}, and the most recent version is~\cite{Graves05}.
One of the inital proposers of LSTM, Prof. Jurgen Schmidhuber, did a survey on the most common variations
of the model last year~\cite{Greff15}, which is the reference for this short discussion.

A single LSTM neuron is presented in Figure~\ref{fig:lstmneuron}. As it can be seen in the picture,
we still have the recurrent connections from the regular RNNs, but now there are multiple entry points
that control the flow of information through the network. Although omitted from the picture, all the
gates are biased, as is suggested in Equations~\ref{eq:equationsLSTM}. The main components, their
role and relevance, are explained as follows.

\begin{figure}[!t]
	\centering
	\includegraphics[width=2.5in]{figures/lstmneuron.png}
    \caption{A complete LSTM neuron, with all the features as described in~\cite{Graves05}. Source:~\cite{Greff15}}
	\label{fig:lstmneuron}
\end{figure}

\begin{itemize}
    \item \textbf{Input Gate} -- this is the input gate, where the relative importance of each feature of
the input vector at time $t$, $\mb{x}^{(t)}$, and the output vector at the previous time step, $\mb{y}^{(t-1)}$,
are weighed in, producing an output $\mb{i}^{(t)}$.

    \item \textbf{Block Input Gate} -- as the name implies, this gate controls the flow of information
from the input gate to the memory cell. It also receives the input vector and the previous output
set, producing an output $\mb{z}^{(t)}$. The \textbf{activation function of this gate can be either},
but the most common choice is the \textbf{hyperbolic hangent}.

    \item \textbf{Forget Gate} -- its role is to control the contents of the Memory Cell, either to
set or reset them, using the \textit{Hadamard Elementwise} vector multiplication of its output at time
$t$, $\mb{c}^{(t-1)}$. The activation function of this gate is \textbf{always sigmoid}, and the resulting
signal is $\mb{f}^{(t)}$.

@@

    \item \textbf{Memory Cell} -- the cornerstone of the LSTM neuron. This is the memory element of
the neuron, where the previous state is kept, and updated accordingly to the dynamics of the gates
that connect to it. Also, this is where the peephole connections come from.

    \item \textbf{Output Activation} -- the output of the Memory Cell goes through this activation
function that, as the gate activation function, can be any, but the \textbf{hyperbolic tangent}
is the most common choice.

    \item \textbf{Peepholes} -- direct connections \textit{from} the memory cell that allow for gates
to `peep' at the states of the memory cell. They were added after the initial 1997 formulation, and their
absence was proven to have a minimal performance impact~\cite{Greff15}. For this reason, they were ommitted
in this architecture.

@@

The operation of each set of gates of the layer is given by the following set of equations
\begin{eqnarray}
    \mb{z}^{(t)} & = & g(\mb{W}_z \mb{x}^{(t)} + \mb{R}_z \mb{y}^{(t-1)} + \mb{b}_z) \nonumber\\
    \mb{i}^{(t)} & = & \sigma(\mb{W}_i \mb{x}^{(t)} + \mb{R}_i \mb{y}^{(t-1)} + \mb{p}_i \odot \mb{c}^{(t-1)} + \mb{b}_i) \nonumber\\
    \mb{f}^{(t)} & = & \sigma(\mb{W}_f \mb{x}^{(t)} + \mb{R}_f \mb{y}^{(t-1)} + \mb{p}_f \odot \mb{c}^{(t-1)} + \mb{b}_f) \nonumber\\
    \mb{o}^{(t)} & = & \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \nonumber\\
    \mb{c}^{(t)} & = & \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \nonumber \\
    \mb{y}^{(t)} & = & \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) \label{eq:equationsLSTM}
\end{eqnarray}
where $\odot$ is the Hadamard multiplication. The $i$-th element of the previous vectors in bold corresponds
to the value of the $i$-th neuron in the layer, which is a very convenient and compact representation of the
whole layer. Furthermore, if the layer has $N$ LSTM neurons and $M$ inputs (i.e. the size of the layer that
precedes this), we see that the input weight matrices $\mb{W}_*$ have size $N \times M$, and the  recurrent
weight matrices $\mb{R}_*$ are square matrices of size $N \times N$, and that the bias weight matrices $\mb{b}_*$
and the matrices $\mb{y}^{(t)}$ through $\mb{c}^{(t)}$ have size $N \times 1$, and are thus, in fact, vectors.

\section{Related Work}\label{sec:relwork}
LSTM Networks are nowadays one of the state of the art algorithms in deep-learning, and their performance is
superior to that of other kinds of RNNs and Hidden Markov Models, both of which are generally used to address
the same set of problems where LSTM are employed, namely predicting and producing classification decisions
from time-series data. A very comprehensive description of applications can be found in one of the initial
authors webpage dedicated to the subject~\footnote{http://people.idsia.ch/~juergen/rnn.html}.

The uses of LSTM Networks are plentiful: in \textbf{Handwriting Recognition}~\cite{ICDAR09}, where they surpass
HMM-based models in optical character recognition~\cite{Breuel13}; in \textbf{Speech Recognition}, where, for
instance, Graves et al.~\cite{Graves13}, in 2013, set a new record on the TIMT Phoneme Recognition Benchmark,
and also in \textbf{Large Scale Acoustic Modelling of Speech}~\cite{Sak14}. Other uses include
\textbf{Handwriting Synthesis}~\cite{Graves13_2}, \textbf{Translation}~\cite{Sustkever14}, \textbf{Biomedical Applications} such
as protein homology detection~\cite{Hochreiter07}, \textbf{Music Analysis and Composition}, such as the transcription of piano music
to MIDI~\cite{Bock12} and automated composition~\cite{Coca13} and improvisation~\cite{Eck02}, and lastly \textbf{Video and Image Analysis} as
in~\cite{Vinyals14},~\cite{Donahue14} and~\cite{Donahue14_2}.

Hitherto, there is but one actual implementation of an LSTM network in hardware, published recently
(March 2016) by Chang et al.~\cite{Chang15}. A 2 layer LSTM network (two layers in series) was implemented,
with 128 neurons each, which, for processing 1000 samples, yielded an execution time of $0.932 \, \si{\second}$. Assuming
that both layers are equal, and thus have the same execution time, this yields an approximate execution time of
$29.13 \, \si{\micro\second}$ per incoming sample (dividing the total execution time by $2 \times 1000$).
Therefore, this network is able to perform around 34.3 thousands of forward propagations per second.
Therefore ts performance is well below this work, which sits at \textbf{0.487 millions} of forward propagations per second.
This is because the work in~\cite{Chang15} does not have a full level of parallelism when compared to the proposed design of
Section~\ref{sec:proparch}, and it makes use of external memory to store the weights. This last point is understandable,
since the authors use the FPGA core as a co-processor for the main CPU, and not as a standalone implementation.

@@

\subsection{Activation Function Calculation}\label{sec:proprarch_af}
In order to evaluate the transcendental activation functions $\sigma(x)$ and $\tanh(x)$, \textbf{Polynomial Approximations} were used, as detailed in~\cite{Muller05},
since evaluating a polynomial does not have high memory usage needs (as opposed to Table Methods, for instance) and, if the polynomial degree is sufficiently low, the
number of multiplications needed is low enough to not pose a restriction both on resources (now DSP slices, and not memory) and in speed (number of number of clock cycles
needed to output a result).

The error minimization strategy used to find the optimal polynomial was the \textbf{Least Maximum Approximation}, where
the \emph{maximum} error is \emph{minimized}, making use of Remez's Algorithm, which sets a system of $n+2$ linear equations such as Equation~\ref{eq:remezline}

\begin{eqnarray}\label{eq:remezline}
    & p(x_i) - f(x_i) = (-1)^{n+1} \epsilon \Leftrightarrow  \nonumber \\
		& p_0 + p_1 x_i + p_2 x_i^2 + \cdots + p_n x_i^n - f(x_i) = (-1)^{n+1} \epsilon
\end{eqnarray}
where $i \in \left[0, n+1\right]$. Of course, both functions have horizontal assymptotes, which can be used as the value in the edges,
but instead of performing the optimization in the single interval in between the chosen points from where the module evaluates the function
as the value of the assymptote, the interval was further split in 4, in order to have lower degree polynomials -- simpler to evaluate -- at a reduced error cost.
The algorithm was run using Python, and targeted second degree polynomials for each interval. After that, the Verilog model was described, where the correct
coefficients are loaded according to the interval where the incoming operand is located. The evaluation of the polynomial
was accomplished using the \textbf{Horner's Rule}, noting that

\begin{equation}\label{eq:factorPol}
p(x) = p_0 + p_1x + p_2x^2 = p_0 + x(p_1 + xp_2)
\end{equation}
and, therefore, the calculation is simply the procedure of multiplying the operand by a value and adding a constant to the result,
repeated two times. Again the internal machine state controls which values are multiplied and added according to the pipeline state.
The output of the Sigmoid HDL block  and the $\tanh$ HDL block are presented in Figures~\ref{fig:nonlin-out-sig}
and~\ref{fig:nonlin-out-tanh}, respectively. Due to the internal datapath pipeline, the module takes 5 clock cycles to output a result.

\begin{figure}[!t]
    \centering
    \includegraphics[width=\linewidth]{figures/nonlin-out.eps}
    \caption{Plot of the output of the Sigmoid Calculator HDL module}
    \label{fig:nonlin-out-sig}
\end{figure}

\begin{figure}[!t]
    \centering
    \includegraphics[width=\linewidth]{figures/nonlin-out-tanh.eps}
    \caption{Plot of the output of the Hyperbolic Tangent HDL module}
    \label{fig:nonlin-out-tanh}
\end{figure}

\subsection{Matrix-vector Dot Product Calculation}\label{sec:proprarch_dot}
From Equations~\ref{eq:equationsLSTM}, we see that the weight matrices $\mb{W}_*$ and $\mb{R}_*$ are
multiplied by the input vector $\mb{x}$ and the layer output vector $\mb{y}$, respectively. This way,
this block implements matrix-vector multiplication to perform those calculations, and is parameterized in
order to accommodate networks of various sizes, since $\mb{W}_*$ has size $N\times M$, and $\mb{R}_*$
has size $N\times N$ -- $x$ has length $M$ (the number of inputs to the layer), while $y$ has length $N$
(the number of neurons in the layer). This way, if a layer with different parameters is needed, either in terms of
number of inputs or number of neurons, we only need to change the respective parameter before the synthesis stage,
instead of having to redesign the whole block for that particular size.

@@

\begin{algorithm}
\begin{algorithmic}
\For {$i$ = $1:N$}
    \For {$j$ = $1:M$}
    \State $y_i := y_i + A_{ij} \cdot x_j$
    \EndFor
\EndFor
\end{algorithmic}
\caption{Matrix-vector multiplication of a matrix}
\label{alg:matvec-alg}
\end{algorithm}
This operation has a computational complexity of $O(n^2)$. Each of the $i$-th components of the output vector $y$ can be calculated \textbf{in parallel},
each only requiring the corresponding $i$-th line from the matrix. Following this approach, matrix-vector multiplication can now be performed in \textbf{linear time},
which is one of the great advantages of custom-tailored hardware solutions.

Although this solution only requires one multiplication per row of the input matrix (i.e. $N$ multiplications), if the row
size is large, we may run out of resources in the FPGA; therefore, some sort of \textit{resource multiplexing} strategy must
be used to ensure the flexibility of the solution to accomodate networks of larger dimensions. The solution found for this
design was to \emph{share} the multiplication slice among the rows of the matrix: in a direct implementation of Algorithm~\ref{matvec-alg},
each multiplication slice was responsible for producing the $i$-th element of the output vector $y$ (of size $N$), therefore the final
result for the vector would be ready in $M$ clock cycles (i.e. the number of columns); now, defining a
parameter $K_G = \frac{\text{Number of rows}}{\text{Number of multipliers}}$ -- the number of rows that share the same multiplier -- the
same multiplier is responsible for producing several $i$-th elements of the output vector, in consecutive time slots of $M$ clock cycles.
For instance, in an $8\times2$ matrix scenario, with $K_G = 2$, we would have 4 multipliers, and the output vector
elements $y_0$, $y_2$, $y_4$ and $y_6$ would be ready after $M=2$ clock cycles, and the remaining -- $y_1$, $y_3$, $y_5$ and $y_7$ --  are
ready after another two clock cycles, that is $2M = 4$ clock cycles after the calculations began.

Figures-\ref{fig:mem-arrayprod} and~\ref{fig:array-prod} depict a diagram of the memory access for the Matrix, and the row multiplication
units within the module, respectively, where $K_G = 4$ and for the same matrix and vector sizes as before. Note that in this situation,
we would only have 2 multipliers, and the module would be composed of two multiplication units, such as those in Figure~\ref{fig:array-prod},
that work in parallel. They address a particular column using the signal \verb+colSel+, which is used by the RAM module to output the corresponding
column of the matrix (in regard to the input vector, obviously this signal selects only a single position), depicted in Figure~\ref{fig:mem-arrayprod}.
The dark shaded part of the memory is used by the first multiplier, and the light shaded is used by the other, in parallel for a
fixed \verb+rowMux+ -- this signal is produced by the control unit of the module, and essentially operates the left multiplexer and right demultiplexer
of Figure~\ref{fig:array-prod}, that allows to choose the proper position of the weight column and to write to the correct output vector position, respectively.
In this example, for \verb+rowMux+=0, the control unit increments \verb+colSel+ from 0 to $M$, and thus evaluating $y_0$ and $y_4$. After this, \verb+rowMux+
is incremented to 1, and the process repeats until \verb+rowMux+ reaches $K_G-1$. Therefore, we have the correct result vector in a time
proportional to $K_G \cdot M$: a 3 clock cycle overhead is added to the previous estimate, since the memory only outputs the appropriate column
in the \emph{next clock cycle}, and that this module is pipelined both at the input and at the output in order to increase the maximum clock frequency.

@@

\subsection{Gate Module}\label{sec:proparch_gate}
The Gate modules are responsible for producing the internal signal vectors for $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$. If we
note that, according to~\cite{Greff15}, the removal of peepholes (the signals $\mb{p}_*$) does not compromise significantly the performance of the network,
we can omit them in order to simplify the Gate Module and reduce the usage of DSP slices. This way, a Gate module needs to perform three tasks

>>>>

%% Additional Packages
\usepackage{cite}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{siunitx}
\usepackage[utf8]{inputenc}

%% Macros
\newcommand{\mb}[1]{\mathbf{#1}}
\renewcommand\IEEEkeywordsname{Keywords}

@@

\begin{abstract}
Our work proposes a hardware architecture for a Long Short-Term Memory (LSTM) Neural Network, aiming
to outperform software implementations, by exploiting its inherent parallelism.
The main design decisions are presented, along with the proposed network architecture. A description of the main
building blocks of the network is also presented. The network is synthesized for various sizes and platforms,
and the performance results are presented and analyzed. Our synthesized network achieves a 251 times speed-up
over a custom-built software network, running on an i7-3770k Desktop computer, proving the benefits of parallel computation
for this kind of network.
\end{abstract}

\begin{IEEEkeywords}
Neural Networks, Long Short-Term, FPGA, Reconfigurable Hardware, Machine Learning
\end{IEEEkeywords}

\section{Introduction}\label{sec:intro}
\IEEEPARstart{N}{eural} Networks are one of the most commonly used techniques in Deep Learning. This
particular type of network, a Long Short-Term Memory (LSTM) Network, is a recursive network, in which the neuron outputs in a certain
time step are also fed as inputs in the next time step, and since it possesses memory, the system can make
sense of patterns within data sequences, unlike classical recursive neural networks.
These algorithms have been profusely implemented in software, and their practical applications are plentiful.
However, the benefits of the inherent parallelism offered by a dedicated hardware
platform are not exploited, and there are relatively few implementations of Machine Learning algorithms in
these kind of platforms.

Hardware platforms could achieve a considerable speed-up over software implementations, which would prove
useful for high data throughput systems, where the calculation overhead is critical and limits the performance.
Furthermore, a hardware implementation can even benefit \emph{offline} Deep Learning tasks by providing the results
of a given experiment faster than in a regular CPU, yielding an increase in scientific productivity.

Hitherto, there is only one implementation, the one of~\cite{Chang15}, but its performance is undermined
by the external memory access. Our implementation aims to make use of internal FPGA memory resources,
and therefore achieve a higher throughput.

Summarizing, this article begins by explaining what are LSTM networks -- and their mathematical
details -- in Section~\ref{sec:lstmnn}, followed by a quick overview of their state
of the art in Section~\ref{sec:relwork}. Section~\ref{sec:proparch} outlines the proposed
hardware architecture and its main constituent modules, and the performance and synthesis
results are reported in Section~\ref{sec:results}. Concluding remarks are given in
Section~\ref{sec:concl}.

\section{LSTM Neural Networks}\label{sec:lstmnn}
LSTM Networks were originally formulated in~\cite{Hoch97}, but their formulation has been incrementally
updated in~\cite{Gers00} and~\cite{Gers2000}, and the most recent version is~\cite{Graves05}.
One of the inital proposers of LSTM, Prof. Jürgen Schmidhüber, did a survey on the most common variations
of the model~\cite{Greff15}, which is the reference for this short discussion.

A single LSTM neuron is presented in Figure~\ref{fig:lstmneuron}. As can be seen in the picture,
we still have the recurrent connections from the regular RNNs, but now there are multiple entry points
that control the flow of information through the network. Although omitted from the picture, all the
gates are biased, as defined in Equations~\ref{eq:equationsLSTM}. The role and relevance of the main components can
be summarized as follows.

\begin{figure}[!t]
	\centering
	\includegraphics[width=1.1\linewidth]{figures/lstmneuron.png}
    \caption{A complete LSTM neuron, with all the features as described in~\cite{Graves05}. Source:~\cite{Greff15}}
	\label{fig:lstmneuron}
\end{figure}

\begin{itemize}
    \item \textbf{Input Gate} -- this is the input gate, where the relative importance of each feature of
the input vector at time $t$, $\mb{x}^{(t)}$, and the output vector at the previous time step, $\mb{y}^{(t-1)}$,
are weighed, producing an output $\mb{i}^{(t)}$.

    \item \textbf{Block Input Gate} -- as the name implies, this gate controls the flow of information
from the input gate to the memory cell. It also receives the input vector and the previous output
set, producing an output $\mb{z}^{(t)}$. The \textbf{activation function of this gate can both} the logistic
sigmoid, $\sigma(x) = 1 \over (1+e^{-x})$, or the hyperbolic tangent, $\tanh(x)$ but the most common choice is the
\textbf{hyperbolic hangent}.

    \item \textbf{Forget Gate} -- its role is to control the contents of the Memory Cell, either to
set or reset them, using the \textit{Hadamard} vector multiplication of its output at time
$t$, $\mb{c}^{(t-1)}$. The activation function of this gate is \textbf{always sigmoid}, and the resulting
signal is $\mb{f}^{(t)}$.

@@

    \item \textbf{Memory Cell} -- the cornerstone of the LSTM neuron. This is the memory element of
the neuron, where the previous state is kept, and updated according to the dynamics of the gates
that connect to it. Also, this is where the peephole connections come from.

    \item \textbf{Output Activation} -- the output of the Memory Cell goes through this activation
function that, as the gate activation function, is the \textbf{hyperbolic tangent}.

    \item \textbf{Peepholes} -- direct connections \textit{from} the memory cell to the gates, that allow them
to `peep' at the states of the memory cell. They were added after the initial 1997 formulation, and their
absence was proven to have a minimal performance impact~\cite{Greff15}. For this reason, they were omitted
in this architecture.

@@

The operation of each set of gates of the layer is given by the following set of equations, where vectors are
represented by bold, lower-case letters, and matrices are bold, upper-case letters
\begin{eqnarray}
    \mb{z}^{(t)} & = & g(\mb{W}_z \mb{x}^{(t)} + \mb{R}_z \mb{y}^{(t-1)} + \mb{b}_z) \nonumber\\
    \mb{i}^{(t)} & = & \sigma(\mb{W}_i \mb{x}^{(t)} + \mb{R}_i \mb{y}^{(t-1)} + \mb{p}_i \odot \mb{c}^{(t-1)} + \mb{b}_i) \nonumber\\
    \mb{f}^{(t)} & = & \sigma(\mb{W}_f \mb{x}^{(t)} + \mb{R}_f \mb{y}^{(t-1)} + \mb{p}_f \odot \mb{c}^{(t-1)} + \mb{b}_f) \nonumber\\
    \mb{o}^{(t)} & = & \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \nonumber\\
    \mb{c}^{(t)} & = & \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \nonumber \\
    \mb{y}^{(t)} & = & \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) \label{eq:equationsLSTM},
\end{eqnarray}
where $\odot$ is the Hadamard multiplication. The $i$-th element of the previous vectors in bold corresponds
to the value of the $i$-th neuron in the layer, which is a very convenient and compact representation of the
whole layer. Furthermore, if the layer has $N$ LSTM neurons and $M$ inputs (i.e. the size of the layer that
precedes this), we see that the input weight matrices $\mb{W}_*$ have size $N \times M$, and the  recurrent
weight matrices $\mb{R}_*$ are square matrices of size $N \times N$, and that the bias weight vectors $\mb{b}_*$
and the vectors $\mb{y}^{(t)}$ through $\mb{c}^{(t)}$ have size $N$.

\section{Related Work}\label{sec:relwork}
LSTM Networks are nowadays one of the state-of-the-art algorithms in Deep Learning, and their performance is
superior to that of other kinds of RNNs and Hidden Markov Models, both of which are generally used to address
the same set of problems where LSTM are employed, namely predicting and producing classification decisions
from time-series data. A very comprehensive description of applications can be found in one of the initial
authors webpage dedicated to the subject~\footnote{http://people.idsia.ch/\~{}juergen/rnn.html}.

The uses of LSTM Networks are plentiful: in \textbf{Handwriting Recognition}~\cite{ICDAR09}, where they surpass
HMM-based models for optical character recognition~\cite{Breuel13}; in \textbf{Speech Recognition}, where, for
instance, Graves et al.~\cite{Graves13}, in 2013, set a new record on the TIMT Phoneme Recognition Benchmark,
and also in \textbf{Large Scale Acoustic Modelling of Speech}~\cite{Sak14}. Other uses include
\textbf{Handwriting Synthesis}~\cite{Graves13_2}, \textbf{Translation}~\cite{Sustkever14}, \textbf{Biomedical Applications} such
as protein homology detection~\cite{Hochreiter07}, \textbf{Music Analysis and Composition}, such as the transcription of piano music
to MIDI~\cite{Bock12} and automated composition~\cite{Coca13} and improvisation~\cite{Eck02}, and lastly \textbf{Video and Image Analysis} as
in~\cite{Vinyals14, Donahue14, Donahue14_2}.

Hitherto, there is but one actual implementation of an LSTM network in hardware, published recently
(March 2016) by Chang et al.~\cite{Chang15}. A 2 layer LSTM network (two layers in series) was implemented in FPGA,
with 128 neurons each, which, for processing 1000 samples, yielded an execution time of \SI{0.932}{\second}. Assuming
that both layers are equal, and thus have the same execution time, this yields an approximate execution time of
$29.13 \, \si{\micro\second}$ per incoming sample (dividing the total execution time by $2 \times 1000$).
Therefore, the network~\cite{Chang15} is able to perform around \num{34.3d3} forward propagations per second.
Although the network is large, its performance is well below this work, which achieves \num{487d3} forward propagations per second, for that network size.
This is because the work in~\cite{Chang15} does not have a full level of parallelism when compared to the proposed design of
Section~\ref{sec:proparch}, and it makes use of external memory to store the weights. On the other hand, the authors use it as a co-processor for the main CPU, and not as a
standalone implementation.

@@

\subsection{Activation Function Calculation}\label{sec:proprarch_af}
In order to evaluate the transcendental activation functions $\sigma(x)$ and $\tanh(x)$, \textbf{Polynomial Approximations} were used, as detailed in~\cite{Muller05},
since evaluating a polynomial does not have high memory usage requirements (as opposed to Table Methods, for instance) and, if the polynomial degree is sufficiently low, the
number of multiplications needed is small enough to not pose a restriction both on resources (now DSP slices, and not memory) and on speed (number of clock cycles
needed to output a result).

The error minimization strategy used to find the optimal polynomial was the \textbf{Least Maximum Approximation}, where
the \emph{maximum} error is \emph{minimized}, making use of Remez's Algorithm, which produces a system of $n+2$ linear equations such as Equation~\ref{eq:remezline}

\begin{eqnarray}\label{eq:remezline}
    & p(x_i) - f(x_i) = (-1)^{n+1} \epsilon \Leftrightarrow  \nonumber \\
		& p_0 + p_1 x_i + p_2 x_i^2 + \cdots + p_n x_i^n - f(x_i) = (-1)^{n+1} \epsilon
\end{eqnarray}
where $i \in \left[0, n+1\right]$. Of course, both functions have horizontal asymptotes, which can be used as the value in the edges.
Instead of performing the optimization in the single interval in between the chosen points from where the module evaluates the function
as the value of the asymptote, the interval was further split in 4, in order to have lower degree polynomials -- simpler to evaluate -- at a reduced error cost.
The algorithm was run using Python, and targeted second degree polynomials for each interval. The coefficients achieved for the Sigmoid and Hyperbolic Tangent functions
are reported in Tables~\ref{tab:coefs-sigm} and ~\ref{tab:coefs-tanh}. The maximum approximation errors were, respectively, \num{1.408d-3} and \num{1.21d-2}.

\begin{table}
	\caption{Polynomial Coefficients for the Sigmoid approximation}
	\label{tab:coefs-sigm}
    \centering
  \begin{tabular}{ | c | c | c | c | }
    \hline
    $p_0$      &  $p_1$     &  $p_2$        &  Interval  \\
		\hline
    0          & 0          &  0             & $x < -6$ \\
    \hline
    0.20323428 & 0.0717631  & 0.00642858     & $-6 \leq x \leq -3$ \\
    \hline
		0.50195831 & 0.27269294 & 0.04059181     & $-3 \leq  x \leq 0$ \\
		\hline
		0.49805785 & 0.27266221 & 0.04058115     &  $0 \leq  x \leq 3$ \\
		\hline
		0.7967568 & 0.07175359 & 0.00642671      & $3 \leq  x \leq 6$ \\
		\hline
		1         &  0         &   0             & $x > 6$ \\
		\hline
  \end{tabular}
\end{table}

\begin{table}
	\caption{Polynomial Coefficients for the $\tanh(x)$ approximation}
	\label{tab:coefs-tanh}
    \centering
  \begin{tabular}{ | c | c | c | c | }
    \hline
    $p_0$      &  $p_1$     &  $p_2$        &  Interval  \\
		\hline
    -1          & 0          &  0             & $x < -3$ \\
    \hline
    -0.39814608 & 0.46527859 &  0.09007576   & $-3 \leq x \leq -1$ \\
    \hline
		0.0031444 & 1.08381219 & 0.31592922      & $-1 \leq  x \leq 0$ \\
		\hline
		-0.00349517 & 1.08538355 & -0.31676793   &  $0 \leq  x \leq 1$ \\
		\hline
		0.39878032 & 0.46509003 & 0.09013554     & $1 \leq  x \leq 3$ \\
		\hline
		1         &  0         &   0             & $x > 3$ \\
		\hline
  \end{tabular}
\end{table}
The Verilog model was described, where the correct
coefficients are loaded according to the interval where the incoming operand is located. The evaluation of the polynomial
was accomplished using the \textbf{Horner's Rule}, that is

\begin{equation}\label{eq:factorPol}
p(x) = p_0 + p_1x + p_2x^2 = p_0 + x(p_1 + xp_2)
\end{equation}
and, therefore, the calculation is simply the procedure of multiplying the operand by a value and adding a constant to the result,
repeated two times. Again the internal machine state controls which values are multiplied and added according to the pipeline state.
Due to the internal datapath pipeline, each module takes 5 clock cycles to output one result.

\subsection{Matrix-vector Dot Product Calculation}\label{sec:proprarch_dot}
From the Set of Equations~\ref{eq:equationsLSTM}, we see that the weight matrices $\mb{W}_*$ and $\mb{R}_*$ are
multiplied by the input vector $\mb{x}$ and the layer output vector $\mb{y}$, respectively. This block implements
matrix-vector multiplication to perform those calculations, and its description is parameterized in
order to accommodate networks of various sizes, since $\mb{W}_*$ has size $N\times M$, and $\mb{R}_*$
has size $N\times N$, because $x$ has length $M$ (the number of inputs to the layer), while $y$ has length $N$
(the number of neurons in the layer). If a layer with different parameters is needed, either in terms of
number of inputs or number of neurons, we only need to change the respective parameter before the synthesis stage,
instead of having to redesign the whole block for that particular size.

@@

\begin{algorithm}
\begin{algorithmic}
\For {$i$ = $1:N$}
    \For {$j$ = $1:M$}
    \State $y_i := y_i + A_{ij} \cdot x_j$
    \EndFor
\EndFor
\end{algorithmic}
\caption{Matrix-vector multiplication of a matrix}
\label{alg:matvec-alg}
\end{algorithm}
This operation has a computational complexity of $O(n^2)$. Each of the $i$-th components of the output vector $y$ can be calculated \textbf{in parallel},
each one only requiring the corresponding $i$-th line from the matrix. Using this approach, matrix-vector multiplication can now be performed in \textbf{linear time},
which is one of the great advantages of custom-tailored hardware solutions.

Although this solution only requires one multiplication per row of the input matrix (i.e. $N$ multiplications), if the row
size is large, we may run out of resources in the FPGA; therefore, some sort of \textit{resource multiplexing} strategy must
be used to ensure the flexibility of the solution to accommodate networks of larger dimensions. The solution found for this
design was to \emph{share} the multiplication slice among the rows of the matrix: in a direct implementation of Algorithm~\ref{alg:matvec-alg},
each multiplication slice is responsible for producing the $i$-th element of the output vector $y$ (of size $N$), therefore the final
result for the vector would be ready in $M$ clock cycles (i.e. the number of columns); now, defining a
parameter $K_G$, such that

\begin{equation}\label{eq:kg}
K_G = \frac{\text{Number of rows}}{\text{Number of multipliers}}.
\end{equation}
This parameters represents the number of rows that share the same multiplier. The
same multiplier is responsible for producing several $i$-th elements of the output vector, in consecutive time slots of $M$ clock cycles.
For instance, in an $8\times2$ matrix scenario, with $K_G = 2$, we would have 4 multipliers, and the output vector
elements $y_0$, $y_2$, $y_4$ and $y_6$ would be ready after $M=2$ clock cycles, and the remaining -- $y_1$, $y_3$, $y_5$ and $y_7$ --  are
ready after another two clock cycles, that is $2M = 4$ clock cycles after the calculations began.

Figures~\ref{fig:mem-arrayprod} and~\ref{fig:array-prod} depict a diagram of the memory access for the Matrix, and the row multiplication
units within the module, respectively, where $K_G = 4$ and for the same matrix and vector sizes as before. Note that in this situation,
we would only have 2 multipliers, and the module would be composed of two multiplication units, such as those in Figure~\ref{fig:array-prod},
that work in parallel. They address a particular column using the signal \verb+colSel+, which is used by the RAM module to output the corresponding
column of the matrix (in regard to the input vector, obviously this signal selects only a single position), depicted in Figure~\ref{fig:mem-arrayprod}.
The dark shaded part of the memory is used by the first multiplier, and the light shaded one is used by the other, in parallel for a
fixed \verb+rowMux+. This signal is produced by the control unit of the module, and essentially operates the left multiplexer and right demultiplexer
of Figure~\ref{fig:array-prod}, that allows choosing the proper position of the weight column and writing to the correct output vector position, respectively.
In this example, for \verb+rowMux+=0, the control unit increments \verb+colSel+ from 0 to $M$, and thus evaluates $y_0$ and $y_4$. After this, \verb+rowMux+
is incremented to 1, and the process repeats until \verb+rowMux+ reaches $K_G-1$. Therefore, we have the correct result vector in a time
proportional to $K_G \times M$: a 3 clock cycle overhead is added to the previous estimate, since the memory only outputs the appropriate column
in the \emph{next clock cycle}, and this module is pipelined both at the input and at the output, in order to increase the maximum clock frequency.

@@

\subsection{Gate Module}\label{sec:proparch_gate}
The Gate modules are responsible for producing the internal signal vectors for $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$. This way,
each Gate module needs to perform three tasks
